Wikipedia Text Mining and Surprisal Analysis

### 2019-07-18 surprisal/MacWhinney poster/poster_plotting.Rmd
matched up wals features and wiki slopes

# pdf("surprisals.pdf", width = 10, height = 5.5)
data %>%
  filter(length == 8, language != "Mandarin") %>%
   ggplot(aes(x = word_order, y = s,
                        color = speaker_role,
                        label = speaker_role)) +
  facet_grid(language~context, scales = "free") +
  geom_point(size = 1.5) +
  geom_line(size = 1.5) +
  scale_color_brewer(palette = "Set1") + 
  theme(legend.position = "none") +
  geom_text(data = label_data, size = 6) + 
  labs(x = "Word Position", y = "Average Information")
# dev.off()
```

>>>>

# pdf("surprisals.pdf", width = 10, height = 5.5)
data %>%
  filter(length == 8, language != "Mandarin") %>%
   ggplot(aes(x = word_order, y = s,
                        color = speaker_role,
                        label = speaker_role)) +
  facet_grid(context~language, scales = "free") +
  geom_point(size = 1.5) +
  geom_line(size = 1.5) +
  scale_color_brewer(palette = "Set1") + 
  theme(legend.position = "none") +
  geom_text(data = label_data, size = 6) + 
  labs(x = "Word Position", y = "Average Information")
# dev.off()
```

```{r plotting features}
models <- read_feather(here("Wikipedia/models.feather"))

sf <- models %>% 
  filter(p.value == 0, estimate > 0) %>% 
  select(name) %>% 
  rename("Some Important Features" = name) %>% 
  slice(c(1, 10, 12, 15, 23, 29))

sf$Examples <- c("Noun-relative clause; relative clause-noun", 
                 "No cases; 6 cases", 
                 "SOV; VSO", 
                 "Only singular; singular and plural", 
                 "Affix; distinct word", 
                 "Prefixes; suffixes")

sf %>% grid.table()

```

### 2019-07-18 surprisal/WALS_features.Rmd
matched up wals features and wiki slopes

```{r setup, include=FALSE}
library(tidyverse)
library(lingtypology)
library(feather)
library(here)
library(missMDA)
library(lsa)
library(reshape2)
library(broom)
knitr::opts_chunk$set(echo = TRUE)
```

```{r read in data}
wals <- read_csv(here("Data/wals_isos.csv"))

@@

wiki <- read_csv(here("Data/relative_ngrams.csv"))
LANGUAGES <- wiki %>% pull(language) %>% unique()
```

```{r see which WALS features are the most complete for our languages}
all_languages <- wals %>% 
  filter(rowSums(is.na(.)) < 130) %>%
  group_by(language) %>%
  slice(1) %>%
  ungroup() 
```

@@

common_languages <- common_df %>%
  pull(language)
```

@@

```{r cosines}
unigram_cosines <- wiki %>%
  filter(language %in% common_languages, gram == "Unigram") %>%
  select(-gram) %>% 
  column_to_rownames("language") %>%
  t() %>%
  cosine() %>%
  as_data_frame(rownames = "language1") %>%
  gather(language2, unigram_cosine, -language1) %>%
  mutate(unigram_cosine = 1 - abs(unigram_cosine))

@@

all_distances %>% 
  ggplot(aes(x = feature)) +
    geom_smooth(aes(y = unigram_cosine, color = "#0033CC"), 
                method = 'lm', se = F) + 
    geom_smooth(aes(y = bigram_cosine, color = "#6699FF"), 
                method = 'lm', se = F) +
    geom_smooth(aes(y = trigram_cosine, color = "#33CCFF"), 
                method = 'lm', se = F) +
    geom_point(aes(y = unigram_cosine, color = "#0033CC")) + 
    geom_point(aes(y = bigram_cosine, color = "#6699FF")) + 
    geom_point(aes(y = trigram_cosine, color = "#33CCFF")) + 
    ylab("Cosine Space Distance") + 
    xlab("Feature Space Distance") + 
    scale_color_manual(name = "Gram", 
                       values = c("#0033CC", "#6699FF", "#33CCFF"), 
                       labels = c("Unigrams", "Bigrams", "Trigrams")) + 
    ggtitle("Cosine space distance regressed on feature spaced distance")
```

@@

wf_pairwise <- expand.grid(language1 = wf_sub$language,
                           language2 = wf_sub$language) %>%
  left_join(wf_sub, by = c("language1" = "language")) %>%
  rename(value1 = value) %>%
  left_join(wf_sub, by = c("language2" = "language", "feature")) %>%
  rename(value2 = value) %>%
  mutate(same = value1 == value2) %>%
  select(-value1, -value2) %>%
  left_join(cosines, by = c("language1", "language2"))

@@

```{r cleaning}
wiki <- read_feather(here("Data/relative_ngrams.feather"))
wiki %>% 
  group_by(language) %>% 
  mutate(order = 1:n()) %>% 
  ungroup() %>% 
  filter(order <= 15) %>% 
  select(-order) %>%
  select(language, gram, estimate, cut) %>% 
  spread(cut, estimate) %>% 
  rename(Slope1 =`1`, Slope2 =`2`, Slope3 =`3`, Slope4 =`4`, Slope5 =`5`) %>%
  write_csv(here("Data/relative_ngrams.csv"))
```

>>>>

```{r setup, include=FALSE}
library(tidyverse)
library(lingtypology)
library(feather)
library(here)
library(missMDA)
library(lsa)
library(reshape2)
library(broom)
library(janitor)
knitr::opts_chunk$set(echo = TRUE)
```

```{r read in data}
# wals <- read_csv(here("Data/wals_isos.csv"))
wals <- read_csv(here("Data/from_website.csv"))

@@

codes <- read_csv(here("Wikipedia/isos.csv"))
wiki <- read_csv(here("Data/relative_ngrams.csv"))
LANGUAGES <- wiki %>% pull(language) %>% unique()
```

```{r get only languages with enough features}
all_languages <- wals %>% 
  mutate(missing = rowSums(is.na(.))) %>% 
  group_by(language) %>% 
  mutate(minimum = min(missing)) %>%
  filter(missing == minimum) %>% 
  select(-missing, minimum) %>%
  ungroup() 
```

@@

# common_languages <- common_df %>%
#   pull(language)
```

@@

```{r cosines}
common_languages <- read_csv(here("Data/imputed_wals.csv")) %>% 
  pull(language)

unigram_cosines <- wiki %>%
  filter(language %in% common_languages, gram == "Unigram") %>%
  select(-gram) %>% 
  column_to_rownames("language") %>%
  t() %>%
  cosine() %>%
  as_data_frame(rownames = "language1") %>%
  gather(language2, unigram_cosine, -language1) %>%
  mutate(unigram_cosine = 1 - abs(unigram_cosine))

@@

all_distances %>% 
  select(-bigram_cosine) %>% 
  ggplot(aes(x = feature)) +
    geom_smooth(aes(y = unigram_cosine, color = "#0033CC"), 
                method = 'lm', se = F) + 
    # geom_smooth(aes(y = bigram_cosine, color = "#6699FF"), 
    #             method = 'lm', se = F) +
    geom_smooth(aes(y = trigram_cosine, color = "#33CCFF"), 
                method = 'lm', se = F) +
    geom_point(aes(y = unigram_cosine, color = "#0033CC")) + 
    # geom_point(aes(y = bigram_cosine, color = "#6699FF")) + 
    geom_point(aes(y = trigram_cosine, color = "#33CCFF")) + 
    ylab("Information Curve Difference") + 
    xlab("Difference in WALS Features") + 
    scale_color_manual(name = "Gram", 
                       values = c("#0033CC", "#33CCFF"), 
                       labels = c("Unigrams", "Trigrams")) 
```

@@

sub_langs <- wf_sub %>%
  pull(language) %>%
  unique()

wf_pairwise <- expand.grid(language1 = sub_langs,
                           language2 = sub_langs) %>%
  filter(language1 != language2)

compute_sim <- function(language1, language2) {

  wf_sub %>%
    filter(language == language1) %>%
    rename(value1 = value) %>% 
    left_join(wf_sub %>% filter(language == language2) %>% select(feature, value), 
              by = "feature") %>%
    mutate(same = value == value1) %>%
    ungroup() %>%
    group_by(language) %>%
    summarise(feature = sum(same)) %>%
    mutate(language2 = language2) %>%
    rename(language1 = language)
}

feature_space <- map_dfr(1:nrow(wf_pairwise), ~compute_sim(wf_pairwise[.x,"language1"],
                                          wf_pairwise[.x,"language2"]))

@@

```{r get values from website}
languages <- read_csv(here("wals_dataset/languages.csv"))

values <- read_csv(here("wals_dataset/values.csv"))

values %>% 
  left_join(languages, by = c("Language_ID" = "ID")) %>% 
  select(Language_ID, Parameter_ID, Value, Name, Glottocode, ISO639P3code) %>% 
  rename(wals_code = Language_ID, feature = Parameter_ID, value = Value, 
         language = Name, glottocode = Glottocode, iso = ISO639P3code) %>% 
  spread(feature, value)

wf %>% write_csv(here("Data/from_website.csv"))
```

```{r pairing up languages with wikipedia features}
wals <- read_csv(here("Data/from_website.csv"))

all_data <- wiki %>% 
  left_join(codes, by = "language") %>% 
  left_join(wals %>% select(-language), by = "iso") %>% 
  mutate(n = rowSums(is.na(.))) %>% 
  group_by(language) %>% 
  filter(n == min(n)) %>% 
  group_by(language, gram) %>%
  slice(1) %>% 
  ungroup()

all_data %>% 
  write_csv(here("Data/wiki_wals.csv"))
```

```{r cleaning, eval = F}
wiki <- read_feather(here("Data/relative_ngrams.feather"))
wiki %>% 
  group_by(language) %>% 
  mutate(order = 1:n()) %>% 
  ungroup() %>% 
  filter(order <= 15) %>% 
  select(-order) %>%
  select(language, gram, estimate, cut) %>% 
  spread(cut, estimate) %>% 
  rename(Slope1 =`1`, Slope2 =`2`, Slope3 =`3`, Slope4 =`4`, Slope5 =`5`) %>%
  write_csv(here("Data/relative_ngrams.csv"))
```

### 2019-07-18 surprisal/wiki_surprisal.Rmd
matched up wals features and wiki slopes

```{r make dendrogram}
wiki_unigrams <- read.csv(here("Data/relative_ngrams.csv")) %>%
  filter(gram == "Unigram") %>%
  sample_n(100) %>%
  column_to_rownames("language")

@@

wals %>%
  mutate(iso = iso.gltc(glottocode)) %>%
  select(-glottocode) %>% 
  inner_join(codes, by = "iso") %>% 
  select(-language.x) %>%
  rename(language = language.y) %>%
  write_csv(here("Data/wals_isos.csv"))

@@

```{r setup}
wiki_isos <- read_csv(here("Wikipedia/wiki_isos.csv")) %>%
  clean_names() %>%
  select(iso_language_name, x639_1, x639_3) %>% 
  mutate(x639_3 = str_extract(x639_3, "\\w+")) %>% 
  rename(language = iso_language_name) %>% 
  filter(language %in% LANGUAGES)

wiki_langs <- fromJSON(here("Wikipedia/language_dict.json")) %>%
  as_tibble() %>%
  gather(language, x639_1) %>%
  filter(language %in% LANGUAGES) %>%
  left_join(wiki_isos)

@@

>>>>

```{r make dendrogram}
wiki_unigrams <- read.csv(here("Data/relative_ngrams.csv")) %>%
  filter(gram == "Unigram") %>%
  column_to_rownames("language")

@@

wals %>%
  mutate(iso = iso.gltc(glottocode)) %>%
  select(-glottocode) %>% 
  select(language, iso) %>% View()

  inner_join(codes, by = "iso") %>% 
  select(-language.x) %>%
  rename(language = language.y) %>%
  write_csv(here("Data/wals_isos.csv"))

@@

```{r setup}
wiki_isos <- read_csv(here("Wikipedia/wiki_isos.csv")) %>%
  clean_names() %>%
  select(iso_language_name, x639_1, x639_3) %>% 
  # mutate(x639_3 = str_extract(x639_3, "\\w+")) %>% 
  rename(language = iso_language_name) %>% 
  filter(language %in% LANGUAGES)

wiki_langs <- fromJSON(here("Wikipedia/language_dict.json")) %>%
  as_tibble() %>%
  gather(language, x639_1) %>%
  filter(language %in% LANGUAGES) %>%
  write_csv(here("wiki_to_iso.csv"))

%>%
  left_join(wiki_isos)

@@

```{r plotting grams across languages, eval = F}
# how much do the slopes differ from 0? 
wiki %>% 
  group_by(language, gram) %>% 
  summarise(diff = sum(abs(Slope1:Slope5))) %>%
  ggplot(aes(x = gram, y = diff)) + 
  geom_point()
```
